---
title: "FishErIes Size and functional TYpe model (FEISTY) in R "
author: "Ken H Andersen and Karline Soetaert Yixin Zhao"
date: " 2024"
output:
  pdf_document: default
  html_document: default
  word_document: default
vignette: >
  %\VignetteIndexEntry{FishErIes Size and functional TYpe model in R (FEISTY)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require(FEISTY)) devtools::install_github("Kenhasteandersen/FEISTY@R-Package")
library(FEISTY)
palette("ggplot2")  # ggplot2-style palette
```

# Introduction 
The R package **FEISTY** is to document the FEITY model and for further model development and research.
This document describes how to run the FEISTY model in R.

DESCRIBE IT BETTER

# existing parameter data sets

Several functions to create suitable parameter inputs are included:

* setupBasic creates a basic three-species setup as described in Petrik et al (2019) 
* setupBasic2 creates a basic three-species setup as setupBasic(), but generalised to:
more realistic sizes, Generalized size-based feeding, the possiblity of more then 3 size groups in each group
* setupVertical makes a basic four-species setup that distinguishes between visual and twilight predators and that 
includes vertical distribution of zooplankton
* setupPelagicSpecies makes a basic setup with just pelagic fish, and with feeding preferences according to the size ratios of predators and prey.

blabla

## Example 

The model is first run with the default 3-functional group prameters, created with *setupBasic()*. 
Then two similar parameter datasets are created, with rates doubled or halved.

```{r}
p <- setupBasic()
knitr::kable(p$resources, digits=2)
knitr::kable(p$fishes, digits=2)
knitr::kable(p$groups, digits=2)
```

Function *paramAddPhysiology* is used to change the allometric rates:

```{r}
p2 <- paramAddPhysiology(p, ac = 40, am = 8, ae=140)

p3 <- paramAddPhysiology(p, ac = 10, am = 2, ae=35)
```

The model can now be run for all parameter sets; the reulst for the last 20 years are shown.

```{r, fig.width=8, fig.height=8}
out1 <- simulateFEISTY(p=p,  times=seq(0, 200, length.out=1000),bCust=T)
out2 <- simulateFEISTY(p=p2, times=seq(0, 200, length.out=1000),bCust=T)
out3 <- simulateFEISTY(p=p3, times=seq(0, 200, length.out=1000),bCust=T)
# plot(out1, out2, out3, which=5:12, lty=1, lwd=2, subset=time>180)
# plot(out1, out2, out3, which=c("smallZoo", "largeZoo", "smallBenthos", 
#   "totBiomass.smallPel", "totBiomass.largePel",  "totBiomass.Demersals"), 
#   lty=1, lwd=2, subset=time>180)
```



\newpage

# Appendix 1. The size-based model

## Fish dynamics

For each fish stage, the dynamics of its biomass reads (see de Roos et al., 2008):

$$\frac{dC_i}{dt}= G_i - F_i + (e_{a_i} - m_i) \cdot C_i - r_{p_i} \cdot C_i$$
where C_i is expressed e.g. in $gWW/m2$.

The losses due to growth are:
$$F_i = \frac{g_i-m_i}{1-\frac{1}{z_i}^{1-m_i/g_i}}\cdot C_i$$
and the growth in each stage is:
$$G_1=\sum_{i=1}^{N}\{ \psi_{repro} \cdot r_{p_i} \cdot C_i \}$$ 
$$G_i=F_{i-1} \quad for~i>1$$ 

Here $g=(1-\psi_{mat})\cdot e^+$ is the energy available for growth, $e_a^+ = max(0, e_a)$,
and $z_i = \frac{s_{i+1}}{s_{i}}$ is the size ratio of the stages. 

Available energy for growth or reproduction comes from assimilated food (left term) minus basal respiration (right term):
$$e_a  = \psi_{Ass} \cdot \frac{E}{c_{max} + E} \cdot c_{max} - \mu$$
$c_{max}$ is the maximum consumption rate, and the encounter rates, $E$, are calculated as:
$$E=\nu \cdot (\theta \times C)$$
with $\nu$ the clearance rate, and $\theta_{i,j}$ the feeding preference matrix for consumer $i$ feeding on prey $j$.

Reproduction ($r_p$) only occurs when there is an energy surplus:
$$r_p = \psi_{mat}\cdot e^+$$
Total mortality is the sum of predation mortality, basal mortality and fishing-induced mortality:
$$m = \theta^t \times \{ \frac{c_{max}*\nu}{c_{max}+E}\cdot C \} +m_0 +m_F$$

The maximal consumption rate ($c_{max_i}$), clearance rate (or encounter rate, $\nu_i$) and metabolism rate ($\mu_i$) is, for each fish stage estimated as a function of its mean size ($m_i$):

$$c_{max_i} = a_c \cdot m_i^{b_c} ; \quad \nu_i = a_e \cdot m_i^{b_e};\quad \mu_i=a_{\mu} \cdot m_i^{b_{\mu}}$$

## Resource dynamics

Logistic growth or chemostat-like dilution describes how the resource density evolves over time:

$$\frac{dR}{dt}=r \cdot R \cdot (1-\frac{R}{K}) - m_R \cdot R,$$

$$\frac{dR}{dt}=r \cdot (K-R)-m_R \cdot R,$$
where the second term in each equation is the resource mortality due to feeding by fishes. 

# References

R Core Team (2021). R: A language and environment for statistical computing. R Foundation for Statistical
  Computing, Vienna, Austria. URL https://www.R-project.org/.
  
  Van Denderen et al. 2020. Emergent global biogeography of marine fish food webs. Global Ecology and Biogeography, DOI: 10.1111/geb.13348

Petrik, CM, Stock, CA, Andersen, KH, van Denderen, PD, Watson, JR 2019. Bottom-up drivers of global patterns of demersal, forage, and pelagic fishes. Progress in Oceanography, 176, 102124. DOI: 10.1016/j.pocean.2019.102124

De Roos, A.M., Schellekens, T., Van Kooten, T., Van De Wolfshaar, K., Claessen, D., Persson, L., 2008. Simplifying a physiologically structured population model to a stage-structured biomass model. Theor. Popul Biol. 73, 47–62.

