---
title: "GOTM-FEISTY Workflow"
author: "Yixin Zhao"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This vignette demonstrates how to run GOTM-FEISTY simulations and visualize results.

# Prerequisites

- GOTM compiled with FABM-FEISTY support
- FEISTY R package

```{r}
library(FEISTY)
```

# 1. Run GOTM Simulation

## Mac / Linux

```{r}
sim_dir <- "/path/to/simulation_folder"
setwd(sim_dir)
system("./gotm")
```

## Windows

### Option A: Both GOTM and R in WSL (or both in Windows)

If GOTM and R are in the same environment, run directly:

```{r}
sim_dir <- "/path/to/simulation_folder"
setwd(sim_dir)
system("./gotm")
```

### Option B: GOTM in WSL, R in Windows

Run GOTM in WSL terminal first:

```bash
cd /path/to/simulation_folder
./gotm
```

Then in R (Windows), access the output via WSL path:

```{r}
sim_dir <- "//wsl.localhost/Ubuntu/path/to/simulation_folder"
setwd(sim_dir)
```

# 2. Extract Results

```{r}
sim <- ncToSim("output.nc")

# For multi-location files, specify coordinates:
# sim <- ncToSim("output.nc", lon = -120, lat = 35)
```

# 3. Visualize

```{r}
plotBiomasstime(sim)
plotNetwork(sim)
```

# Simulation Object Fields

- `sim$t` - Time vector (years)
- `sim$totBiomass` - Total biomass per functional type
- `sim$u` - Full state matrix (resources + fish by size class)
- `sim$p` - Parameters from setupVertical2()
